---
title: "data clean_jy2965"
author: "Jun Yin"
date: "11/22/2019"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(plotly)
library(readxl)
library(patchwork)
library(rvest)
library(httr)
knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
theme_set(theme_minimal() + theme(legend.position = "bottom"))

```



## Importing Data

```{r, results='hide',message=FALSE}
squrriel = read_csv("./data/2018_Central_Park_Squirrel_Census_-_Squirrel_Data.csv") %>% 
janitor::clean_names()  
skimr::skim(squrriel) # observe the distribution of our dataset
```




## Data Cleaning

reclassify some variables with missing values.
give character variables different level.
Create new variable to represent research date for futher work.



```{r,message=FALSE}
squrriel = read_csv("./data/2018_Central_Park_Squirrel_Census_-_Squirrel_Data.csv") %>% 
janitor::clean_names() %>% 
mutate(age_group = ifelse(age %in% "Adult","Adult",ifelse(age %in% "Juvenile","Juvenile","Unknown")),
      age_group = fct_relevel(as.factor(age_group),c("Juvenile","Adult","Unknown")),
      location = fct_relevel(as.factor(location),c("Above Ground","Ground Plane")),
      primary_fur_color = fct_relevel(as.factor(primary_fur_color),c("Cinnamon","Gray","Black"))) %>% 
separate(date,into=c('day','year'),sep=4) %>% 
separate(day,into = c('month','Date'), sep=2) %>% 
mutate(Day = recode(Date, "06"="Day1", "07"="Day2", "08"="Day3", "10"="Day4", "12"="Day5", "13"="Day6", "14"="Day7", "17"="Day8", "18"="Day9", "19"="Day10", "20"="Day11"),
       Day = fct_relevel(Day, c("Day1", "Day2", "Day3", "Day4", "Day5", "Day6", "Day7", "Day8", "Day9", "Day10", "Day11")))
```

## Descriptive Statistics

```{r}
squrriel %>% 
  mutate(Day = recode(Day, "Day1"=1, "Day2"=2, "Day3"=3, "Day4"=4, "Day5"=5, "Day6"=6, "Day7"=7, "Day8"=8, "Day9"=9, "Day10"=10, "Day11"=11)) %>% 
  group_by(Day, age_group) %>% 
  summarise(
    running = sum(running == "TRUE"),
    chasing = sum(chasing == "TRUE"),
    climbing = sum(climbing == "TRUE"),
    eating = sum(eating == "TRUE"),
    approaches = sum(approaches == "TRUE"),
    foraging = sum(foraging == "TRUE"),
    indifferent = sum(indifferent == "TRUE"),
    kuks = sum(kuks == "TRUE"),
    quaas = sum(quaas == "TRUE"),
    runs_from = sum(runs_from == "TRUE"),
    tail_flags = sum(tail_flags == "TRUE"),
    tail_twitches = sum( tail_twitches== "TRUE")
    ) %>% 
    pivot_longer(
    running:tail_twitches,
    names_to = "activity", 
    values_to = "number_of_ture")%>%
  mutate(activity = as.factor(activity)) %>% 
  ggplot(aes(x=Day, y=number_of_ture, group=activity))+
geom_line(aes(color=activity))+
  geom_point(aes(color=activity))+
  facet_grid(.~age_group)+
  labs(
    title = "Different activity frequency across the experiment",
    x= "Days of experiment",
    y = "Frequency"
    )
  
```

*This graph shows the number of different activities across the period of experiment in each age group.*
*We can see the decreased trend of number of each activity at the end of experiment*


Then we reclassified the activity based on if they were interacted with people. And generated a bar graph describing their frequency.


```{r}
squrriel %>% 
  group_by(Day) %>% 
  summarise(
    running = sum(running == "TRUE"),
    chasing = sum(chasing == "TRUE"),
    climbing = sum(climbing == "TRUE"),
    eating = sum(eating == "TRUE"),
    approaches = sum(approaches == "TRUE"),
    foraging = sum(foraging == "TRUE"),
    indifferent = sum(indifferent == "TRUE"),
    kuks = sum(kuks == "TRUE"),
    quaas = sum(quaas == "TRUE"),
    runs_from = sum(runs_from == "TRUE"),
    tail_flags = sum(tail_flags == "TRUE"),
    tail_twitches = sum( tail_twitches== "TRUE")
    )  %>%  
mutate(noninteraction = running + chasing + climbing + eating + foraging,
         interaction = kuks + quaas + tail_flags + tail_twitches + indifferent + runs_from) %>% 
  select(Day,noninteraction, interaction) %>% 
 plot_ly(x = ~Day, y = ~interaction, type = 'bar', name = 'Interaction', marker = list(color = 'rgb(149,130,189)')) %>%
  add_trace(y = ~noninteraction, name = 'Noninteraction', marker = list(color = 'rgb(100,100,100)')) %>%
  layout(xaxis = list(title = "", tickangle = -45),
         yaxis = list(title = ""),
         margin = list(b = 200),
         barmode = 'group')
```

