---
title: "The Squirrel Project"
output: html_document
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE, message =FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rvest)
library(plotly)
library(ggplot2)
library(ggpubr)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 6,
  out.width = "90%"
)

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

theme_set(theme_minimal() + theme(legend.position = "bottom"))
```


## *Motivation*

Squirrels have always served as a needed source of entertainment in Central Park. Other rodents' population has been

well traced and documented by humans, but not always Squirrels. However, the necessity of mapping the total population

of squirrels in Central Park is observed as it depicts this community's size and growth, which is highly related to the

evaluation of this community's functioning as a source of entertainment, as well as their potential of harboring and 

delivering pathogens to humans.

In this project, we use The Squirrel Census data from NYC Open Data host, which focuses on Eastern gray species

specifically. For this project, we aim at depicting a full picture of the current squirrels community in Central Park, 

with regards to their demographics statistics as well as their living status. We also intend to assess their performance 

and activity of Squirrels in Central Park, by evaluating their interactions with humans.


We believe this project can provide evidence-based information for the Central Park Squirrel community, as well as 

contribute to better evaluation and regulation of the Squirrel population in Central Park. This project also has the 

potential in providing further management insights for the planning and designing aspect of the Central Park.


## *Related Work*


A recent news article published on The New York Times sheds some light on the motivation behind the two-week long Central

Park Squirrel Census. Despite the ubiquity of the Eastern gray, not much was known about it that we, as humans, should 

have figured out from a while ago. The value of such census was not just to delineate a specific behaviour pattern of the 

Eastern gray, but to draw a big picture of the squirrels, the park, and us. We were inspired by the behind-scene 

motivation of such study and aimed to create this website that has interactive features to learn more about the 

distribution of the squirrel population and factors that may affect their behaviour patterns. 


We believe such interactive visualization allows squirrel-lovers to pick a place to better spot the squirrels and to allow 
Central Park’ management team to build a better environment for the squirrel community. 


## *Initial Questions*

In this project, we aimed to examine factors associated with the distribution of the squirrel population in Central Park

as well as factors that may be important in predicting certain squirrel behaviours and squirrel-human/predator 

interactions. Our last aim was to predict and visualize different behaviour pattern of the squirrels. We therefore 

proposed the following research questions: 

* How does squirrel fur color change throughout different hectares in Central park?

* What factors affect squirrel distribution in Central Park? If any association exists, how can we visualize it? 

* How can we predict and visualize squirrel’s behaviour and/or their interaction with humans/predators based on ambient environment and temperature?


## *Data*

1. 2018 Central Park Squirrel Census - Squirrel Data:
(https://data.cityofnewyork.us/Environment/2018-Central-Park-Squirrel-Census-Squirrel-Data/vfnx-vebw)

  This table contains squirrel data for each of the 3,023 sightings, including location coordinates, age, primary and

  secondary fur color, elevation, activities, communications, and interactions between squirrels and with humans.

2. 2018 Central Park Squirrel Census - Hectare Data:
(https://data.cityofnewyork.us/Environment/2018-Central-Park-Squirrel-Census-Hectare-Data/ej9h-v6g2)

This table contains environmental data related to each of the 350 “countable” hectares of Central Park. Examples include 

weather, litter, animals sighted, and human density.

# *Exploratory analysis*

## Load in Dataset

We downloaded the data from the web and called it `hectare`. 

```{r, echo= FALSE, message=FALSE}
hectare = read_csv("./data/2018_Central_Park_Squirrel_Census_-_Hectare_Data.csv") %>% 
janitor::clean_names()
```


## Tidy the Dataset 

We created following new variables: 

`temperature_fehrenheit` : temperatures for different time during a day. 

`rate`: the number of squirrels in each hectare observed by sighters per minute. 

`temperature_category` : there are five categories for temperatures, “below 50”, “50-60”, “60-70”, “above 70”, “others”. 

`noninteraction`: total squirrels'activity count for running, chasing, climbing, eating and foraging.

`interaction` : total activity count for interaction between squirrels and predators or human, including

kukking, quaaing, tail flags, tail twitching, being indifferent and running away. 


```{r, echo=FALSE, message=FALSE}
numextract <- function(string){ 
  str_extract(string, "\\-*\\d+\\.*\\d*")
}

hectare = 
  hectare %>% 
  mutate(
      temperature_fahrenheit = numextract(sighter_observed_weather_data),
      rate = number_of_squirrels/total_time_of_sighting
)

squrriel = read_csv("./data/2018_Central_Park_Squirrel_Census_-_Squirrel_Data.csv") %>% 
janitor::clean_names() %>% 
mutate(age_group = ifelse(age %in% "Adult","Adult",ifelse(age %in% "Juvenile","Juvenile","Unknown")),
      age_group = fct_relevel(as.factor(age_group),c("Juvenile","Adult","Unknown")),
      location = fct_relevel(as.factor(location),c("Above Ground","Ground Plane")),
      primary_fur_color = fct_relevel(as.factor(primary_fur_color),c("Cinnamon","Gray","Black"))) %>% 
separate(date,into=c('day','year'),sep=4) %>% 
separate(day,into = c('month','Date'), sep=2) %>% 
mutate(Day = recode(Date, "06"="Day1", "07"="Day2", "08"="Day3", "10"="Day4", "12"="Day5", "13"="Day6", "14"="Day7", "17"="Day8", "18"="Day9", "19"="Day10", "20"="Day11"),
       Day = fct_relevel(Day, c("Day1", "Day2", "Day3", "Day4", "Day5", "Day6", "Day7", "Day8", "Day9", "Day10", "Day11")))
```

## *Figures*

### Boxplot for Temperature and Rate of Squirrel observed

The boxplot described the relationship between temperature and rate of squirrel observed


```{r, echo=FALSE}
temperature_df =   
hectare %>%   
  filter(temperature_fahrenheit > 29) %>% 
  mutate(
    temperature_fahrenheit = recode(temperature_fahrenheit, `-50` = "50"),
    temperature_category = ifelse(temperature_fahrenheit <= "50", "below 50",
                           ifelse("50"< temperature_fahrenheit & temperature_fahrenheit<= "60", "50-60",
                           ifelse("60"< temperature_fahrenheit & temperature_fahrenheit<= "70", "60-70",
                           ifelse("70"< temperature_fahrenheit, "above 70","others")))),
    temperature_category = as.factor(temperature_category)
 ) %>%
  mutate(
    temperature_category = factor(temperature_category, levels = c("below 50","50-60", "60-70", "above 70"))
  ) %>% 
  plot_ly( y = ~rate, color = ~temperature_category, type = "box", colors = "Set1")

temperature_df

```

During the study period (10/7/2018-10/19/2018), the temperatures' range is between 50 and 70 degrees Fahrenheit.Below 50 

degrees Fahrenheit, the number of squirrels observed drastically declined. Above 50, the change on rate of squirrel 

observed is not significant. 


### Bubble plot for rate of squirrel observed 

The bubble plot describes how conditions affect the rate of squirrel with different levels of litter amount

```{r, echo = FALSE}
condition_litter_dataset = 
  hectare %>% 
  select(hectare_conditions, litter, rate) %>% 
  na.omit() 


condition_litter_dataset = 
condition_litter_dataset %>% 
  mutate(
 average =  ifelse(litter == "Abundant"& hectare_conditions == "Busy","abundant_busy",
                ifelse(litter == "Abundant"& hectare_conditions == "Calm","abundant_calm",
                  ifelse(litter == "Abundant"& hectare_conditions == "Medium","abundant_medium",
         ifelse(litter == "Abundant"& hectare_conditions == "Moderate","abundant_moderate",
            ifelse(litter == "Some"& hectare_conditions == "Busy","some_busy",
                ifelse(litter == "Some"& hectare_conditions == "Calm","some_calm",
                  ifelse(litter == "Some"& hectare_conditions == "Medium","some_medium",
         ifelse(litter == "Some"& hectare_conditions == "Moderate","some_moderate",

ifelse(litter == "None"& hectare_conditions == "Busy","none_busy",
                ifelse(litter == "None"& hectare_conditions == "Calm","none_calm",
                  ifelse(litter == "None"& hectare_conditions == "Medium", "none_medium",
         ifelse(litter == "None"& hectare_conditions == "Moderate","none_moderate", "others"))))))))))))
)

rate_mean = 
condition_litter_dataset %>% 
  group_by(average) %>% 
  select(hectare_conditions, litter, rate, average) %>% 
  mutate(
    rate_average = mean(rate, na.rm = TRUE)
  ) %>% 
  distinct(rate_average) %>% 
  separate(average, into = c("litter", "hectare_condition"), sep = "_") %>% 
  na.omit()

plot = 
rate_mean %>% 
  mutate(
    hectare_condition = factor(hectare_condition, levels = c("calm", "moderate", "medium","busy")),
    litter = factor(litter, levels = c("none", "some", "abundant"))
    ) %>% 
ggballoonplot(x = "hectare_condition", y = "litter", size = "rate_average", fill = "rate_average", alpha = 0.5) + 
  scale_fill_viridis_c(option = "C") +
  labs(
    title = "Litter, Condition and Mean Rate of Squirrel Observed",
    fill = "Mean Rate of Squirrel",
    size = "Mean Rate of Squirrel"
  )

plot
```


The amount of litter and noise levels in the environment did not affect squirrel distribution much; however, when there 

was an abundant amount of litters and high level of noise in the environment, the number of squirrels at the lowest.   

This partially aligns with our expectations that squirrels stay away from the crowds and occupied areas but why other 

levels of litter and noise did not affect the number of squirrels as much worths further investigation.  


### Line Graph for Each Activity Counts

This graph indicates the total activity frequency for juvenile and adult squirrels during study period. The activity 

includes non-interative and interative with predators or human. 

```{r, echo = FALSE}
squrriel %>% 
  mutate(Day = recode(Day, "Day1"=1, "Day2"=2, "Day3"=3, "Day4"=4, "Day5"=5, "Day6"=6, "Day7"=7, "Day8"=8, "Day9"=9, "Day10"=10, "Day11"=11)) %>% 
  group_by(Day, age_group) %>% 
  summarise(
    running = sum(running == "TRUE"),
    chasing = sum(chasing == "TRUE"),
    climbing = sum(climbing == "TRUE"),
    eating = sum(eating == "TRUE"),
    approaches = sum(approaches == "TRUE"),
    foraging = sum(foraging == "TRUE"),
    indifferent = sum(indifferent == "TRUE"),
    kuks = sum(kuks == "TRUE"),
    quaas = sum(quaas == "TRUE"),
    runs_from = sum(runs_from == "TRUE"),
    tail_flags = sum(tail_flags == "TRUE"),
    tail_twitches = sum( tail_twitches== "TRUE")
    ) %>% 
    pivot_longer(
    running:tail_twitches,
    names_to = "activity", 
    values_to = "number_of_ture")%>%
  mutate(activity = as.factor(activity)) %>% 
  ggplot(aes(x=Day, y=number_of_ture, group=activity))+
geom_line(aes(color=activity))+
  geom_point(aes(color=activity))+
  facet_grid(.~age_group)+
  labs(
    title = "Different activity frequency across the experiment",
    x= "Days of experiment",
    y = "Frequency"
    )
  
```

Generally, the activity frequency is significantly higher among adult squirrels than juvenile. 


### Bar Graph for Comparison Between Interative and Non-interative Activity Counts

  This graphs show the frequency of squirrel's non-interative and interative activities with predactors  or human in the park.


```{r, echo = FALSE}
squrriel %>% 
  group_by(Day) %>% 
  summarise(
    running = sum(running == "TRUE"),
    chasing = sum(chasing == "TRUE"),
    climbing = sum(climbing == "TRUE"),
    eating = sum(eating == "TRUE"),
    approaches = sum(approaches == "TRUE"),
    foraging = sum(foraging == "TRUE"),
    indifferent = sum(indifferent == "TRUE"),
    kuks = sum(kuks == "TRUE"),
    quaas = sum(quaas == "TRUE"),
    runs_from = sum(runs_from == "TRUE"),
    tail_flags = sum(tail_flags == "TRUE"),
    tail_twitches = sum( tail_twitches== "TRUE")
    )  %>%  
mutate(noninteraction = running + chasing + climbing + eating + foraging,
         interaction = kuks + quaas + tail_flags + tail_twitches + indifferent + runs_from) %>% 
  select(Day,noninteraction, interaction) %>% 
 plot_ly(x = ~Day, y = ~interaction, type = 'bar', name = 'Interaction', marker = list(color = 'rgb(149,130,189)')) %>%
  add_trace(y = ~noninteraction, name = 'Noninteraction', marker = list(color = 'rgb(100,100,100)')) %>%
  layout(xaxis = list(title = "", tickangle = -45),
         yaxis = list(title = ""),
         margin = list(b = 200),
         barmode = 'group')

```

Overall, non-interative activity counts were higher than interactive activity counts for the whole study period. There 

was an obvious decreasing trend on both activity counts during the last few days. 

## *Discussion* 

### Finding

It is also interesting to see how various factors including temperature, park condition, influence both squirrel’s own 

behaviours as well as their interactions with the external environment (including humans and potential predators). 

A few limitations of this dataset worth mentioning here. The overall sample size was small making stratified analysis 

difficult. We therefore did not perform analysis on specific squirrel activity and could not comment on how the predictors 
of interest (temperature, condition,  ) here will independently affect each activity of interest. In addition, all data 

was only collected for 11 days over a period of two consecutive weeks and the temperatures did not vary as much making our 
conclusion on the association between temperature and squirrel distribution less convincing. A followup problem from this 

is that we were not able to make any inference about change in squirrel’s behaviours based on several predictors tested as 
our samples were limited and were best represent the squirrel population in Central Park. To our knowledge, there was only 
one other squirrel census conducted previously in Atlanta, our analysis based on data collected in 2018 provides the most 

up-to-date information on squirrel behaviors and their interactions with the environment, humans, and other predators. We therefore believe that our attempt to discover factors associated with squirrel distribution or squirrel behaviours still gave some interesting findings. However, to draw a more definitive conclusion, further investigation on significant predictors of squirrel behaviour patterns should be conducted on wider squirrel populations that span wider geographic areas with longer observation period.  




$$ Rate \ of \ Squirrel \ Observed \sim \beta_0 + \beta_1 temperature + \beta_2 litterNone + \beta_3 litterSome + \beta_4 litterAbundant + \beta_5 hectare_conditionsCalm +\beta_6 hectare_conditionsModerate +\beta_7 hectare_conditionsMedium +\beta_8 hectare_conditionsBusy + \varepsilon_i,$$
```{r, echo=FALSE}
hectare_model =
  hectare %>%
  mutate(
    temperature_fahrenheit = as.numeric(temperature_fahrenheit)
  )
linearMod <- lm(rate ~ temperature_fahrenheit+litter+hectare_conditions, data = hectare_model)


linearMod %>%
broom::tidy() %>% 
  select(term, estimate, p.value) %>%
  knitr::kable(digit = 3)
```






